---
title: "Heavy-tailed Gompertz Models with Stan"
author: Sean Anderson
output: html_document
---

Let's fit a Gompertz model with heavy-tailed process error to a time series of
grey heron counts from England. This is population ID 20579 from the Global
Population Dynamics Database.

(Data from Brooks, B., Traill, L. W., Bradshaw, C. J. A. 2006 Minimum viable
population sizes and global extinction risk are unrelated. Ecology Letters, 9
(4), 375-382.)

We'll start by loading some packages and setting some options:

```{r, message=FALSE}
knitr::opts_knit$set(root.dir = "..")
library(rstan)
library(bayesplot)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
library(ggplot2)
theme_set(theme_light())
```

Let's read in the data:

```{r}
d <- read.csv("data/grey-heron.csv")[,2:3]
# d <- read.csv("data/wind-speed.csv")
# d <- read.csv("data/cod-productivity.csv")
names(d) <- c("year", "index")
ggplot(d, aes(year, log(index))) +
  geom_point()
```

You can see the basic model we will work with in the file `R/stan/gompertz.stan`. 

We can fit that model with Stan:

```{r, message=FALSE, warning=FALSE}
m <- stan("R/stan/gompertz.stan",
  data = list(N = nrow(d), y = log(d$index), nu_rate = 0.01),
  iter = 800)
```

Let's look at the parameters from our model:

```{r}
pars <- c("lambda", "b", "sigma_proc", "nu")
print(m, pars = pars)
traceplot(m, pars = pars)
mcmc_hist(as.matrix(m), pars = pars)
# mcmc_areas(as.matrix(m), pars = pars)
# shinystan::launch_shinystan(m)
```

We can extract the posterior samples with the function `rstan::extract()`. Part
of the Stan code calculated predictions from our model. Alternatively, we could
calculate them in R ourselves. We will also calculate residuals from our model.

```{r}
e <- extract(m)
pred <- data.frame(
  est = apply(e$pred, 2, median),
  l = apply(e$pred, 2, quantile, probs = 0.025),
  h = apply(e$pred, 2, quantile, probs = 0.975),
  year = d$year)
pred$resid <- log(d$index) - pred$est
thresh <- qnorm(0.0001, 0, sd = median(e$sigma_proc))
```

Let's plot our predictions with the data overlaid:

```{r}
ggplot(pred, aes(year, est, ymin = l, ymax = h)) +
  geom_ribbon(alpha = 0.4) +
  geom_line() +
  geom_point(aes(y = log(d$index)))
```

And we can look at which residuals exceed some arbitrary threshold:

```{r}
ggplot(pred, aes(year, resid)) +
  geom_line() +
  geom_hline(yintercept = c(thresh, -thresh), lty = 2)
```

As a comparison, we could fit the model with normally distributed process
error. We can compare the models with the leave-one-out information criterion
(LOOIC).

Also notice that the parameter estimates differ.

```{r, message=FALSE, warning=FALSE}
m2 <- stan("R/stan/gompertz-normal.stan",
  data = list(N = nrow(d), y = log(d$index)),
  iter = 800)
```

```{r}
print(m, pars = c("lambda", "b", "sigma_proc", "nu"))
print(m2, pars = c("lambda", "b", "sigma_proc"))

library(loo)
loo(extract_log_lik(m))
loo(extract_log_lik(m2))
```

Let's try fitting our model to a second data set where there aren't extremes.
This time we will use a data set of population counts for house wrens in the
mid 20th century in Illinois, USA. GPDD population ID 28.

(Data reference: Kendeigh, S.C. 1982 Bird populations in east central Illinois:
fluctuations, variation and development over a half-century. Illinois
Biological Monographs, 52:1-136)

```{r}
d_hw <- read.csv("data/house-wren.csv")[,2:3]
names(d_hw) <- c("year", "index")
ggplot(d_hw, aes(year, index)) +
  geom_point()
```

```{r, message=FALSE, warning=FALSE}
m_hw <- stan("R/stan/gompertz.stan", 
  data = list(N = nrow(d_hw), y = log(d_hw$index), nu_rate = 0.01),
  iter = 800)
```

```{r}
print(m_hw, pars = pars)
```

So in this case our estimate of nu (the degrees of freedom parameter) is much
higher. In fact, it approximately matches the prior because the data are not
informative about heavy tails. In this case it reverts to be effectively
normal:

```{r}
prior <- rexp(1e6, rate = 0.01)
prior <- prior[prior > 2]
hist(prior)
median(prior)
abline(v = median(prior), col = "red", lwd = 2)
```

This choice of prior is based on Fernandez C, Steel MFJ (1998) On Bayesian
modeling of fat tails and skewness. J. Am. Stat. Assoc. 93(441):359–371.

Since our models are fit in a Bayesian framework, one useful value to calculate
might be the probability density below some threshold value for the degrees of
freedom parameter. For example:

```{r}
nu <- extract(m, pars = "nu")[[1]]
nu_hw <- extract(m_hw, pars = "nu")[[1]]

sum(nu < 10)/length(nu) # or mean(nu < 10)
sum(nu_hw < 10)/length(nu) 
```

We've often used a value of 10 as a threshold since nu values much above 10
render distributions that are almost indistinguishable from the normal. 

In the first case nu is less than 10 with `r mean(nu < 10)` probability. In the
second case, there is only about a `r mean(nu_hw < 10)` probability of nu < 10.
Remember that even with the prior, there is some probability that nu < 10.

```{r}
mean(prior < 10)
```

An alternative prior, that we have used in our more recent work is 
`gamma(shape = 2, rate = 0.1)` as recommended in 
Juarez, M. A. and M. F. J. Steel. 2010. Model-based clustering of
non-Gaussian panel data based on skew-t distributions. J. Bus. 
Econ. Stat., 28:52–66.

Let's try a state-space version:

```{r, message=FALSE, warning=FALSE}
m <- stan("R/stan/gompertz-ss.stan",
  data = list(N = nrow(d), y = log(d$index), nu_rate = 0.01, sigma_obs = 0.05),
  iter = 2000, control = list(max_treedepth = 15, adapt_delta = 0.9))

# m <- stan("R/stan/gompertz-ss-est.stan",
#    data = list(N = nrow(d), y = log(d$index), nu_rate = 0.01),
#    iter = 8000, control = list(max_treedepth = 15, adapt_delta = 0.99), chains = 8)
```

```{r}
e <- extract(m)
pred <- data.frame(
  est = apply(e$pred, 2, median),
  l = apply(e$pred, 2, quantile, probs = 0.025),
  h = apply(e$pred, 2, quantile, probs = 0.975),
  year = d$year)
ggplot(pred, aes(year, est, ymin = l, ymax = h)) +
  geom_ribbon(alpha = 0.5) +
  geom_line() +
  geom_point(aes(y = log(d$index)))
```

Look at the parameters from our model:

```{r}
print(m, pars = c("U", "pred", "pred_obs"), include = FALSE)
traceplot(m, pars = c("U", "pred", "pred_obs"), include = FALSE)
```

For many other possible heavy-tailed Stan population models, see <https://github.com/seananderson/heavy-tails/blob/master/analysis/1-compile-models.R>